Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-10-16 manuscript/plos/review_1.tex
added parameter space exploration and others

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses. Indeed, it allows to jointly consider the statistical association between genetic variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high scores of association with a phenotype. More specifically, we examine their use to discover new breast cancer susceptibility genes by interrogating a genome-wide association study (GWAS) data set focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of these six methods with respect to the size of the solution, its predictive power, and the stability and the runtime of the methods. Network methods include prior knowledge in the analysis: they might tolerate selecting a gene with a high P-value, as long as it connects other two, low P-value genes. This alows them to give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, at least two network methods selected 14 neighbors of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find significant overlaps between the genes in five of the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable: the average Pearson correlation across all the methods between the genes selected on different subsamples of the data is 0.45, versus 0.57 simply using Bonferroni correction. To compensate this instability, we produce a stable consensus solution, formed by the most consistently selected genes. This stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than the average gene. The network seems organized around \emph{CUL3}, which encodes a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to address known limitations of GWAS, namely their lack of statistical power and the difficulty of their interpretation. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an elegant explanation: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated so that many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we characterize six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as it is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are strongly functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might help alleviate the aforementioned issues.

@@

In this work, we analyze the application of these six network methods on GWAS data. They do so by using different interpretations of the omnigenic model, providing a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer susceptibility genes. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus solution of predisposition to familial breast cancer.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant network (BH adjusted P\=/value~$<$~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)}~Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors. \textbf{(B)}~VEGAS2 P-values of the genes in the PPIN not selected by any network method (12\,213), and of those selected by 1 (575), 2 (73), or 3 (20) methods. \textbf{(C)}~Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(D)}~Solution networks produced by the different methods.}
  \label{fig:solution_overview}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.7\linewidth]{./figures/figure_3.pdf}
  \caption{ {\bf Consensus solution on GENESIS (Section~\ref{methods:methods}) }. Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset (Section~\ref{methods:bcac}). The latter ones are also colored in pink. All gene names are indicated in \nameref{sfig:consensus-names}.}
  \label{fig:consensus}
\end{figure}

@@

We compared the six network methods in a 5-fold subsampling setting (Section \ref{methods:benchmark}). This allowed us to measure five properties (Fig~\ref{fig:benchmark}): size of the solution; sensitivity and specificity of an L1-penalized logistic regression classifier on the selected SNPs; stability; and computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs (Section~\ref{methods:snp2gene}). The largest solutions came from SConES GI (6\,256.6 SNPs), and dmGWAS (4\,255.0 SNPs). LEAN did not produce any solution in any of the subsamples. To determine whether the selected SNPs could be used for patient classification, we computed the performance of the classifier on the \emph{test dataset} (\nameref{sfig:additional-benchmarks}B). The different classifiers displayed similarly poor sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming only at minimizing prediction error. Of course, although these performances are low, we do not expect to separate cases from controls well using exclusively genetic data \cite{deloscamposComplexTraitPredictionEra2018}.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\linewidth]{./figures/figure_4.pdf}
  \caption{{ \bf Comparison of network-based GWAS methods on GENESIS. } Each method was run 5 times on a random subset containing 80\% of the samples, and tested on the remaining samples (Section \ref{methods:benchmark}). As LEAN did not select any gene, it was excluded from all panels except \textbf{D}. \textbf{(A)}~Number of SNPs selected by each method and number of SNPs in the active set found by the classifier(Section~\ref{methods:classifier}). Points are the average over the 5 runs; the error bars represent the standard error of the mean. A grey diagonal line with slope 1 is added for comparison, indicating the upper bound of the active set (i.e. the number of SNPs in the solution). For reference, the active set of Lasso using all the SNPs included, on average, 154\,117.4 SNPs. \textbf{(C)}~Pairwise Pearson correlations of the solutions produced by different methods. A Pearson correlation of 1 means the two solutions are the same. A Pearson correlation of 0 means that there is no SNP in common between the two solutions. \textbf{(D)}~Runtime of the evaluated methods, by type of network used (PPIN or SNP). For gene-based methods, inverted triangles represent the runtime of the algorithm itself, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics.}
  \label{fig:benchmark}
  \end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\linewidth]{./figures/figure_2.pdf}
  \caption{{\bf Drawbacks encountered when using network methods.} \textbf{(A)}~DmGWAS solution, with the genes colored according to the -log$_{10}$ of their P\=/value. \textbf{(B)}~Centrality degree and -log\textsubscript{10} of the VEGAS2 P\=/value in BCAC for each of the nodes in the PPIN. We highlighted the genes selected by each method, and the ones selected by more than one (``Consensus''). We labeled the three most central genes that were picked by any method. \textbf{(C)}~Number of times a gene was selected by either dmGWAS, heinz, LEAN, or SigMod in 100 rewirings of the PPIN (Section~\ref{methods:pathway_enrichment}) and its centrality degree. \textbf{(D)}~Overlap between the solutions of SConES GS, GM or GI in the different genomic regions. SNPs that were not selected in the studied network, but were selected in another one, are displayed in background color. }
  \label{fig:issues}
  \end{figure}

Another peculiarity of network methods is their relationship to degree centrality. We studied random rewirings of the PPIN while preserving node centrality (Section~\ref{methods:rewiring}). In this setting, network methods favored central genes, as they often connect high scoring nodes (Fig~\ref{fig:issues}B). This is despite the fact that highly central genes often had no association to breast cancer susceptibility (Fig~\ref{fig:issues}C). This was especially the case of SigMod, which selected three highly central, unassociated genes: \emph{COPS5}, \emph{CUL3} and \emph{FN1}. As we showed in Section \ref{results:consensus}, and will show in \ref{results:stable-consensus}, there is evidence in the literature of the contribution of the first two to breast cancer susceptibility. With regards to \emph{FN1}, it encodes a fibronectin, a protein of the extracellular matrix involved in cell adhesion and migration. Overexpression of \emph{FN1} has been observed in breast cancer \cite{Ioachim2002}, and it is negatively correlated with poor prognosis in other cancer types \cite{Yi2016,Sponziello2016}.

@@

@@

Most of the network methods, including the consensus, were highly unstable, raising questions about the reliability of the results. We built a new, stable consensus solution using the genes selected most often across the 30 solutions obtained by running the 6 methods on 5 different splits of the data (Section \ref{methods:benchmark}). Such a network is expected to capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 of the solutions, which corresponded to 1\% of all genes selected at least once. The resulting stability-based consensus was composed of 68 genes (Fig \ref{fig:stable-consensus}). This network shares most of the properties of the consensus: breast cancer susceptibility genes are overrepresented (P\=/value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P\=/values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively); the selected genes are more central than average (P\=/value = 1.1 \texttimes{} 10\textsuperscript{-14}); and a considerable number of nodes (19) are isolated.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.7\linewidth]{./figures/figure_5.pdf}
  \caption{{\bf Stable consensus solution on GENESIS (Section~\ref{results:stable-consensus}).} Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the most central genes (\emph{CUL3}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset (Section~\ref{methods:bcac}). The latter ones are also colored in pink. All gene names are indicated in \nameref{sfig:stable-consensus-names}.}
  \label{fig:stable-consensus}
  \end{figure}

@@

DmGWAS, Heinz, HotNet2, SConES and SigMod all yield interesting solutions, which include (but are not limited to) known breast cancer susceptibility genes. In general, the selected genes and SNPs were more central than average, in accordance with the observation that disease genes are more relatively central \cite{pinero_uncovering_2016}. However, very central nodes are also more likely to be connecting any given random pair of nodes, making them more likely to be selected by network methods (Section \ref{results:drawbacks}). Yet, across this article we show that highly central genes that were selected (\emph{COPS5}, \emph{CUL3} and \emph{FN1}) could plausibly be involved in breast cancer susceptibility. Despite these similarities, the solutions obtained were notably different. At one end of the spectrum, SConES and heinz preferred small, highly associated solutions, at the expense of not shedding much light on the etiology of the disease. On the other end, SigMod and dmGWAS gravitate towards larger, less associated solutions which provide a wide overview of the biological context. While this deepens our understanding of the disease and provides biological hypotheses, they require further analyses, which might deter unexperienced practitioners. For instance, examining the network properties of the genes in the solution (e.g. centrality and neighbors). Those might drive the selection of genes, especially of those with low association scores. HotNet2 balances both approaches at the expense of producing the largest solution: a constellation of many, highly associated, small subnetworks. Additionally, all solutions share two drawbacks. First, they are all equally bad at discriminating cases from controls. Yet, the classification accuracy of network methods is similar to that of a classifier trained on the entire genome, which suggests that cases and controls are difficult to separate in the GENESIS dataset. This may be due to limited statistical power, which reduces the ability to identify relevant SNPs; but in any event, we do not expect to be able to separate people who have or will developed cancer from others on the sole basis of their genomes, ignoring all environmental factors and chance events. Second, all methods are remarkably unstable, yielding different solutions for slightly different inputs. This might partly be caused by the instability of the P\=/values themselves in low statistical power settings \cite{halseyFickleValueGenerates2015}. Hence, heinz's conservative transformation of P\=/values, which favors only the most extreme ones, leads to improved stability. Another source of instability might be the redundancy inherent to biological networks, since they are subject to an evolutionary pressure to avoid single points of failure \cite{wagnerAlternativeRoutesMutational2007}. Hence, biological networks will often have multiple paths connecting two high-scoring nodes.

@@

A crucial step for the gene-based methods is the computation of the gene score. In this work we used VEGAS2 \cite{mishra_vegas2:_2015} due to the flexibility it offers to use user-specified gene annotations. However, it presents known problems (selection of an appropriate percentage of top SNPs, long runtimes and P\=/value precision limited to the number of permutations \cite{nakka_gene_2016}), and other algorithms (e.g. PEGASUS \cite{nakka_gene_2016}, SKAT \cite{ionita-laza_sequence_2013} or COMBAT \cite{wang_combat:_2017}) might have more statistical power.

How to handle linkage disequilibrium (LD) is often a concern among GWAS practitioners. % Another important decision is how to handle LD in a GWAS.
VEGAS2 accounts for LD patterns, and hence an LD pruning step would not impact gene-based network methods, although it would speed up VEGAS2's computation time. With regards to SConES, fewer SNPs would lead to simpler SNP networks and, possibly, shorter runtimes. However, as mentioned in Section~\ref{results:drawbacks}, LD patterns seem to drive SConES' solutions, and an LD pruning step could potentially alter them.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Network methods provide a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses. They discover susceptibility genes by jointly consider the statistical association between genetic variation and a phenotype, measured in a genome-wide association study (GWAS), and the biological context of each gene, represented as a network. Hence, a network method could select a gene with a high P-value of association if it connects multiple low P-value genes in the network. In this work, we studied six network methods which identify subnetworks with high scores of GWAS association with a phenotype. This allows them to give more compelling results than standard SNP- and gene-level GWAS analyses, recovering causal subnetworks tightly related to cancer susceptibility. We applied them to GENESIS, a GWAS on French women with a family history of breast cancer and who tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We critically compared these six methods, discussing the impact of their different mathematical frameworks, and the parameter choices. Additionally, we performed an in-depth benchmarking with respect to the size and predictive power of their solutions, as well as their stability and runtimes. Importantly, we found significant overlaps between the genes in five of the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Since most of the methods produced reasonable solutions, we proposed to combine them into a consensus solution, containing the genes selected by at least two methods. This aggregation brought further insights. For instance, it contained \emph{COPS5}, a gene related to multiple hallmarks of cancer, and 14 of its neighbors. The main drawback of network methods with regards to conventional $\chi^2$-based GWAS was instability i.e. network methods' outputs changed more in front of small perturbations in the input. To compensate for this instability, we proposed a stable consensus solution, formed by the most consistently selected genes in different subsamples of the data. This stable consensus was composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than most genes. The network was organized around \emph{CUL3}, which encodes a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, we showed how network address limitations of GWAS, namely their lack of statistical power and the difficulty of their interpretation. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an appealing explanation: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated so that many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we characterize six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as it is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are unknown. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are strongly inter-related and influence each other's function, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might help alleviate the aforementioned issues.

@@

In this work, we analyze the application of these six network methods on GWAS data. They use different interpretations of the omnigenic model, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer susceptibility genes. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus solution of predisposition to familial breast cancer.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant solution (BH adjusted P\=/value~$<$~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)}~Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors. \textbf{(B)}~VEGAS2 P-values of the genes in the PPIN not selected by any network method (12\,213), and of those selected by 1 (575), 2 (73), or 3 (20) methods. \textbf{(C)}~Solution networks produced by the different methods. \textbf{(D)}~Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes).}
  \label{fig:solution_overview}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_3.pdf}
  \caption{ {\bf Consensus solution on GENESIS (Section~\ref{methods:methods}) }. \textbf{A} ~Manhattan plots of genes; in black, the ones in the consensus solution. The Bonferroni threshold is indicated by a red line (1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{B} Consensus network. Each gene is represented by a pie chart, which shows the methods that selected it. We enlarged the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset (Section~\ref{methods:bcac}). \textbf{C} The nodes are in the same disposition as in panel B, but every gene name is indicated. We colored in pink the names of genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset}
  \label{fig:consensus}
\end{figure}

@@

We compared the six network methods in a 5-fold subsampling setting (Section \ref{methods:benchmark}). This allowed us to measure five properties (Fig~\ref{fig:benchmark}): size of the solution; sensitivity and specificity of an L1-penalized logistic regression classifier on the selected SNPs; stability; and computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs (Section~\ref{methods:snp2gene}). The largest solutions came from SConES GI (6\,256.6 SNPs), and dmGWAS (4\,255.0 SNPs). LEAN did not produce any solution in any of the subsamples. Using different combinations of parameters (Section \ref{methods:parameters}), we computed how good each of the methods was at recovering the results of a conventional GWAS on BCAC (Section \ref{methods:bcac}, Fig~\ref{fig:benchmark}B). SConES exhibits the largest area under the curve, since, when $\lambda = 0$ (network topology is disregarded), it is equivalent to a Bonferroni correction. Between the remaining network methods, they have similar areas under the curve, with heinz being the one with the largest area.

To determine whether the selected SNPs could be used for patient classification, we computed the performance of the classifier on the \emph{test dataset} (\nameref{sfig:additional-benchmarks}B). The different classifiers displayed similarly poor sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming only at minimizing prediction error. Of course, although these performances are low, we do not expect to separate cases from controls well using exclusively genetic data \cite{deloscamposComplexTraitPredictionEra2018}.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_4.pdf}
  \caption{{ \bf Comparison of network-based GWAS methods on GENESIS. } Each method was run 5 times on a random subset containing 80\% of the samples, and tested on the remaining samples (Section \ref{methods:benchmark}). As LEAN did not select any gene, it was excluded from all panels except \textbf{D}. \textbf{(A)}~Number of SNPs selected by each method and number of SNPs in the active set found by the classifier(Section~\ref{methods:classifier}). Points are the average over the 5 runs; the error bars represent the standard error of the mean. A grey diagonal line with slope 1 is added for comparison, indicating the upper bound of the active set (i.e. the number of SNPs in the solution). For reference, the active set of Lasso using all the SNPs included, on average, 154\,117.4 SNPs. \textbf{(B)} True positive rate and true negative rate, using significant SNPs (for SConES and $\chi^2$ + Bonferroni) and genes (for the remaining methods) in BCAC (Section~\ref{methods:bcac}) as true positives, of multiple parameter combinations for different methods (Section \ref{results:parameters}). \textbf{(C)}~Pairwise Pearson correlations of the solutions produced by different methods. A Pearson correlation of 1 means the two solutions are the same. A Pearson correlation of 0 means that there is no SNP in common between the two solutions. \textbf{(D)}~Runtime of the evaluated methods, by type of network used (PPIN or SNP). For gene-based methods, inverted triangles represent the runtime of the algorithm itself, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics.}
  \label{fig:benchmark}
  \end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_2.pdf}
  \caption{{\bf Drawbacks encountered when using network methods.} \textbf{(A)}~DmGWAS solution, with the genes colored according to the -log$_{10}$ of their P\=/value. \textbf{(B)}~Centrality degree and -log\textsubscript{10} of the VEGAS2 P\=/value in BCAC for each of the nodes in the PPIN. We highlighted the genes selected by each method, and the ones selected by more than one (``Consensus''). We labeled the three most central genes that were picked by any method. \textbf{(C)}~Number of times a gene was selected by either dmGWAS, heinz, LEAN, or SigMod in 100 rewirings of the PPIN (Section~\ref{methods:pathway_enrichment}) and its centrality degree. \textbf{(D)}~Overlap between the solutions of SConES GS, GM or GI in the different genomic regions. SNPs that were not selected in the studied network, but were selected in another one, are displayed in background color. }
  \label{fig:issues}
  \end{figure}

Another peculiarity of network methods is their relationship to degree centrality. We studied random rewirings of the PPIN while preserving node centrality (Section~\ref{methods:rewiring}). In this setting, network methods favored central genes, as they often connect high scoring nodes (Fig~\ref{fig:issues}B). This is despite the fact that highly central genes often had no association to breast cancer susceptibility (Fig~\ref{fig:issues}C). This was especially the case of SigMod, which selected three highly central, unassociated genes in both the PPIN and in many of the random rewirings: \emph{COPS5}, \emph{CUL3} and \emph{FN1}. As we showed in Section \ref{results:consensus}, and will show in \ref{results:stable-consensus}, there is evidence in the literature of the contribution of the first two to breast cancer susceptibility. With regards to \emph{FN1}, it encodes a fibronectin, a protein of the extracellular matrix involved in cell adhesion and migration. Overexpression of \emph{FN1} has been observed in breast cancer \cite{Ioachim2002}, and it is negatively correlated with poor prognosis in other cancer types \cite{Yi2016,Sponziello2016}.

@@

\subsection{Different parameters produce similarly-sized solutions}
\label{results:parameters}

We explored the parameter space of the different methods by running them under different combinations of parameters (Section \ref{methods:parameters}). In agreement with their formulations (Section \ref{methods:methods}), larger values of certain parameters produce less astringent solutions (\nameref{sfig:parameters}A): for HotNet2 and heinz, we examined the threshold to decide each gene has a positive score or zero; for dmGWAS, the \texttt{d} parameter controls how far neighbors could be added; node involved how far the search from the explored subnetwork; for SigMod \texttt{nmax} specifies the maximum size of the solution; and for LEAN, it was the P-value threshold to consider a solution significant. Two parameters had the opposite effect (larger is more stringent): SigMod's \texttt{maxjump}, which sets the threshold to consider an increment in $\lambda$ ``large enough''; and SConES $\eta$, where higher values produce more stringent solutions. However, two of the parameters did not have the expected effect: dmGWAS' \texttt{r}, which controls the minimum increment in the score required to add an additional gene; and SigMod's \texttt{maxjump}, which sets the threshold to consider an increment in $\lambda$ ``large enough''. In both cases, the size of the solution was very similar across the different values. Despite the differences in size, the size of the solutions was relatively robust to the choice of parameters (\nameref{sfig:parameters}B).

We computed the Pearson correlation between the different solutions as in Section \ref{methods:algorithm_comparison} to study how the parameters affected which genes and SNPs were selected (\nameref{sfig:parameters}C). This showed that dmGWAS and SigMod are robust to certain parameters: dmGWAS' output was mostly determined by the parameter $d$, rather than $r$; SigMod's by $nmax$, rather than $maxjump$.

SConES presented an interesting case in terms of feature selection: most of the explored combinations of parameters led to trivial solutions (either all the SNPs, or none of them were included) (\nameref{sfig:parameters}A). To explore a more meaningful parameter space, we selected the parameters in two rounds. First, we explored the whole sample space. Then, we focused in a range of $\eta$ and $\lambda$ 1.5 orders of magnitude above and below the best parameters, respectively. This second parameter space was more diverse, and allowed to find more interesting solutions.

@@

Most of the network methods, including the consensus, were highly unstable, raising questions about the reliability of the results. We built a new, \emph{stable consensus} solution using the genes selected most often across the 30 solutions obtained by running the 6 methods on 5 different splits of the data (Section \ref{methods:benchmark}). Such a network is expected to capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 of the solutions, which corresponded to 1\% of all genes selected at least once. The resulting stability-based consensus was composed of 68 genes (Fig \ref{fig:stable-consensus}). This network shares most of the properties of the consensus: breast cancer susceptibility genes are overrepresented (P\=/value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P\=/values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively); the selected genes are more central than average (P\=/value = 1.1 \texttimes{} 10\textsuperscript{-14}); and a considerable number of nodes (19) are isolated.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_5.pdf}
  \caption{ {\bf Stable consensus solution on GENESIS (Section~\ref{results:stable-consensus}) }. \textbf{A} ~Manhattan plots of genes; in black, the ones in the stable consensus solution. The Bonferroni threshold is indicated by a red line (1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{B} Stable consensus network. Each gene is represented by a pie chart, which shows the methods that selected it. We enlarged the most central gene (\emph{CUL3}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset (Section~\ref{methods:bcac}). \textbf{C} The nodes are in the same disposition as in panel B, but every gene name is indicated. We colored in pink the names of genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset}
  \label{fig:stable-consensus}
  \end{figure}

@@

DmGWAS, Heinz, HotNet2, SConES and SigMod all yield interesting solutions, which include (but are not limited to) known breast cancer susceptibility genes. In general, the selected genes and SNPs were more central than average, in accordance with the observation that disease genes are more relatively central \cite{pinero_uncovering_2016}. However, very central nodes are also more likely to be connecting any given random pair of nodes, making them more likely to be selected by network methods (Section \ref{results:drawbacks}). Yet, across this article we show that highly central genes that were selected (\emph{COPS5}, \emph{CUL3} and \emph{FN1}) could plausibly be involved in breast cancer susceptibility. Despite these similarities, the solutions obtained were notably different. At one end of the spectrum, SConES and heinz preferred small, highly associated solutions, at the expense of not shedding much light on the etiology of the disease. On the other end, SigMod and dmGWAS gravitate towards larger, less associated solutions which provide a wide overview of the biological context. While this deepens our understanding of the disease and provides biological hypotheses, they require further analyses. For instance, a user might need to examine the centrality of the selected genes, and discern the extent to which that property is driving the selection of each gene. HotNet2 balances both approaches at the expense of producing the largest solution: a constellation of many, highly associated, small subnetworks. Additionally, all solutions share two drawbacks. First, they are all equally bad at discriminating cases from controls. Yet, the classification accuracy of network methods is similar to that of a classifier trained on the entire genome, which suggests that cases and controls are difficult to separate in the GENESIS dataset. This may be due to limited statistical power, which reduces the ability to identify relevant SNPs; but in any event, we do not expect to be able to separate people who have or will developed cancer from others on the sole basis of their genomes, ignoring all environmental factors and chance events. Second, all methods are remarkably unstable, yielding different solutions for slightly different inputs. This might partly be caused by the instability of the P\=/values themselves in low statistical power settings \cite{halseyFickleValueGenerates2015}. Hence, heinz's conservative transformation of P\=/values, which favors only the most extreme ones, leads to improved stability. Another source of instability might be the redundancy inherent to biological networks, since they are subject to an evolutionary pressure to avoid single points of failure \cite{wagnerAlternativeRoutesMutational2007}. Hence, biological networks will often have multiple paths connecting two high-scoring nodes.

@@

A crucial step for the gene-based methods is the computation of the gene score. In this work we used VEGAS2 \cite{mishra_vegas2:_2015} due to the flexibility it offers to use user-specified gene annotations. However, it presents known problems (selection of an appropriate percentage of top SNPs, long runtimes and P\=/value precision limited to the number of permutations \cite{nakka_gene_2016}), and other algorithms like PEGASUS \cite{nakka_gene_2016}, SKAT \cite{ionita-laza_sequence_2013} or COMBAT \cite{wang_combat:_2017} might have more statistical power.
